Type-Safe Modular Parsing

### 2017-01-14 paper/sections/Inheritance.tex
Section 3

Unfortunately, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to
add new operations to the AST, as usual in the OO style~\cite{wadler1998expression}. On the hand
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a surprise that parser generators that also generate
AST code (such as ANTLR~\cite{parr2013definitive}), normally generate code based on
visitors. Clearly this seems a better choice since generating code
based on the {\sc Interpreter} or {\sc Composite} pattern would
simply be too inflexible, due to the dual limitation on extensibility.

In the remainder of this section we show a technique for
defining \emph{extensible} parsing code. That is new
language constructs can be modularly added without changing
existing parsing code. The technique employs standard OO
style mechanisms such as subtyping and inheritance, as well as,
Packrat parsing combinators.

@@

\subsection{Multiple Extensions and Independent Extensibility}
A nice feature of Scala is its support for a form of multiple
inheritance.  Scala has a linearized-style multiple inheritance for
traits~\cite{odersky2004overview}. This can be very helpful when
composing several languages, and to achieve independent
extensibility~\cite{odersky2005independently}. 
Suppose now we want to compose the parsers for expressions
from pre-defined languages \inlinecode{LanguageA} and
\inlinecode{LanguageB} using alternative. The new parser can be built
by inheriting both parsers at the same time:

@@

As we demonstrated, inheritance with method overriding is the key technique to obtain semantic modularity.
It enables type-safe code reuse and separate compilation for parsing OO style ASTs.

>>>>

Unfortunately, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to
add new operations to the AST, as usual in the OO
style~\cite{wadler1998expression}. On the other hand
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a surprise that parser generators that also generate
AST code (such as ANTLR~\cite{parr2013definitive}), normally generate code based on
visitors. Clearly this seems a better choice since generating code
based on the {\sc Interpreter} or {\sc Composite} pattern would
simply be too inflexible, due to the dual limitation on extensibility.

In the remainder of this section we show a technique for
defining \emph{extensible} parsing code. That is new
language constructs can be modularly added without changing
existing parsing code. The technique employs standard OO
style mechanisms such as subtyping and inheritance, as well as,
Packrat parsing combinators, and enables a better integration bettwen 
parsing and OO ASTs.

@@

\subsection{Multiple Extensions and Independent Extensibility}
A nice feature of Scala is its support for a form of multiple
inheritance.  Scala has a linearized-style mixin inheritance for
traits~\cite{odersky2004overview}. This can be very helpful when
composing several languages, and to achieve independent
extensibility~\cite{odersky2005independently}. 
Suppose now we want to compose the parsers for expressions
from pre-defined languages \inlinecode{LanguageA} and
\inlinecode{LanguageB} using alternative. The new parser can be built
by inheriting both parsers at the same time:

@@

\paragraph{Conflicts and/or ambiguity}

As demonstrated, inheritance with method overriding is the key
technique to obtain semantic modularity.  It enables type-safe code
reuse and separate compilation for parsing OO style ASTs.
